---
title: "EDS232 Lab2b: Community Ordination"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Learning Objectives {.unnumbered}

In this lab, you will play with **unsupervised classification** techniques while working with **ecological community** datasets.

- **Ordination** orders sites near each other based on similarity. It is a multivariate analysis technique used to effectively collapse dependent axes into fewer dimensions, i.e. dimensionality reduction.

  - **Principal Components Analyses (PCA)** is the most common and oldest technique that assumes linear relationships between axes. You will follow a non-ecological example from [Chapter 17 Principal Components Analysis | Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/pca.html) to learn about this commonly used technique.

  - **Non-metric MultiDimensional Scaling (NMDS)** allows for non-linear relationships. This ordination technique is implemented in the R package [`vegan`](https://cran.r-project.org/web/packages/vegan/index.html). You'll use an ecological dataset, species and environment from lichen pastures that reindeer forage upon, with excerpts from the [vegantutor vignette](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf) ([source](https://github.com/jarioksa/vegandocs)) to apply these techniques:
    - **Unconstrained ordination** on species using NMDS;
    - Overlay with environmental gradients; and
    - **Constrained ordination** on species and environmnent using another ordination technique, **canonical correspondence analysis (CCA)**.

## 1 Ordination

**Ordination** orders sites near each other based on similarity. It is a multivariate analysis technique used to effectively collapse dependent axes into fewer dimensions, i.e. dimensionality reduction. It also falls into the class of **unsupervised learning** because a “response term” is not used to fit the model.

## Principal Components Analysis (PCA)

Although this example uses a non-ecological dataset, it goes through the materials walk through the idea and procedure of conducting an ordination using the most widespread technique.

Please read the entirety of [Chapter 17 Principal Components Analysis | Hands-On Machine Learning with R](https://bradleyboehmke.github.io/HOML/pca.html#finding-principal-components). Supporting text is mentioned below where code is run.

### 1.1.1 Prerequisites

See supporting text: [17.1 Prerequisites](https://bradleyboehmke.github.io/HOML/pca.html#prerequisites-14)

```{r}
# load R packages
librarian::shelf(
  dplyr, ggplot2, h2o)

# set seed for reproducible results
set.seed(42)

# get data
url <- "https://koalaverse.github.io/homlr/data/my_basket.csv"
my_basket <- readr::read_csv(url)
dim(my_basket)

my_basket
```

From [Section 1.4](https://bradleyboehmke.github.io/HOML/intro.html#data):

- `my_basket.csv`: Grocery items and quantities purchased. Each observation represents a single basket of goods that were purchased together.
  * Problem type: unsupervised basket analysis
  * response variable: NA
  * features: 42
  * observations: 2,000
  * objective: use attributes of each basket to identify common groupings of items purchased together.

### Performing PCA in R

See supporting text: [17.4 Performing PCA in R](https://bradleyboehmke.github.io/HOML/pca.html#performing-pca-in-r)

```{r}
h2o.no_progress()  # turn off progress bars for brevity
h2o.init(max_mem_size = "5g")  # connect to H2O instance
```

```{r}
# convert data to h2o object
my_basket.h2o <- as.h2o(my_basket)

# run PCA
my_pca <- h2o.prcomp(
  training_frame = my_basket.h2o,
  pca_method = "GramSVD",
  k = ncol(my_basket.h2o),
  transform = "STANDARDIZE",
  impute_missing = TRUE,
  max_runtime_secs = 1000)
my_pca
```

**Question: Why is the pca_method of “GramSVD” chosen over “GLRM”? See HOMLR 17.4 Performing PCA in R.**

**The pca_method of "GramSVD" is chosen over GLRM because our data contains mostly numeric data, and GLRM would be applied if the data was mostly categorical.**

**Question: How many initial principal components are chosen with respect to dimensions of the input data? See HOMLR 17.4 Performing PCA in R.**

**The number of initial principal components that are chosen with respect to dimensions of the input data is the total number of columns in `my_basket.h2o`, which is 42.**

```{r}
my_pca@model$eigenvectors %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, reorder(feature, pc1))) +
  geom_point()
```

**Question: What category of grocery items contribute most to PC1? (These are related because they're bought most often together on a given grocery trip)**

**The `adult beverages` contribute most to PC1**

```{r}
my_pca@model$eigenvectors %>%
  as.data.frame() %>%
  mutate(feature = row.names(.)) %>%
  ggplot(aes(pc1, pc2, label = feature)) +
  geom_text()
```

**Question: What category of grocery items contribute the least to PC1 but positively towards PC2?**

**The category of grocery items that contributes the least to PC1 but positively towards PC2 is healthy fresh foods such carrots, spinach, leeks, etc.**

### Eigenvalue criterion

See supporting text: [17.5.1 Eigenvalue criterion](https://bradleyboehmke.github.io/HOML/pca.html#eigenvalue-criterion).


```{r}
# Compute eigenvalues
eigen <- my_pca@model$importance["Standard deviation", ] %>%
  as.vector() %>%
  .^2

# Sum of all eigenvalues equals number of variables
sum(eigen)
## [1] 42
```

```{r}
# Find PCs where the sum of eigenvalues is greater than or equal to 1
which(eigen >= 1)
```

```{r}
# Extract PVE and CVE
ve <- data.frame(
  PC  = my_pca@model$importance %>% seq_along(),
  PVE = my_pca@model$importance %>% .[2,] %>% unlist(),
  CVE = my_pca@model$importance %>% .[3,] %>% unlist())

# Plot PVE and CVE
ve %>%
  tidyr::gather(metric, variance_explained, -PC) %>%
  ggplot(aes(PC, variance_explained)) +
  geom_point() +
  facet_wrap(~ metric, ncol = 1, scales = "free")
```

**Question: How many principal components would you include to explain 90% of the total variance?**

**35 principal components would explain 90% of the total variance.**

```{r}
# How many PCs required to explain at least 75% of total variability
min(which(ve$CVE >= 0.75))
```

```{r}
# Screee plot criterion
data.frame(
  PC  = my_pca@model$importance %>% seq_along,
  PVE = my_pca@model$importance %>% .[2,] %>% unlist()) %>%
  ggplot(aes(PC, PVE, group = 1, label = PC)) +
  geom_point() +
  geom_line() +
  geom_text(nudge_y = -.002)
```

**Question: How many principal components to include up to the elbow of the PVE, i.e. the “elbow” before plateau of dimensions explaining the least variance?**

**8 principal components are included up to the elbow of the PVE, before the line flattens out.**

**Question: What are a couple of disadvantages to using PCA? See HOMLR 17.6 Final thoughts.**

**Some disadvantages of using PCA are that PCA can be highly affected by outliers and PCA does not perform as well in very high dimensional space where complex nonlinear patterns often exist because the PC directions are linear.**

## 1.2 Non-metric MultiDimensional Scaling (NMDS)

### 1.2.1 Unconstrained Ordination on Species

See supporting text: **2.1 Non-metric Multidimensional scaling** in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf):

```{r}
# load R packages
librarian::shelf(
  vegan, vegan3d)

# vegetation and environment in lichen pastures from Vare et al (1995)
data("varespec") # species
data("varechem") # chemistry

varespec %>% tibble()
```

**Question: What are the dimensions of the varespec data frame and what do rows versus columns represent?**

**The dimensions of the varispec dataframe are 24 rows by 44 columns. Columns describe species with scientific names. Rows represent sites. Using ordination, we will reduce the complexity os all these columns and order things.**

```{r}
vare.dis <- vegdist(varespec)
vare.mds0 <- monoMDS(vare.dis)
stressplot(vare.mds0)
```

**Question: The “stress” in a stressplot represents the difference between the observed inpnut distance versus the fitted ordination distance. How much better is the non-metric (i.e., NMDS) fit versus a linear fit (as with PCA) in terms of \(R^2\)?**

**In terms of \(R^2\, the non-metric fit is better by a small margin, 0.047. This difference in values shows the advantage of using the NMDS approach that fits non-linear relationships instead of using the more traditional ordination approach that assumes a linear fit.**

```{r}
ordiplot(vare.mds0, type = "t")
```
**Question: What two sites are most dissimilar based on species composition for the first component MDS1? And two more most dissimilar sites for the second component MDS2?**

**For the first component MDS1, the sites 28 and 5 are most dissimilar, and for the second component MDS2, the sites 21 and 14 are most dissimilar.**

```{r}
vare.mds <- metaMDS(varespec, trace = FALSE)
vare.mds
```

```{r}
plot(vare.mds, type = "t")
```

**Question: What is the basic difference between metaMDS and monoMDS()? See 2.1 Non-metric Multidimensional scaling of vegantutor.pdf.**

**For non-metric multi-dimensional scaling, metaMDS uses several random starts and selects among similar solutions with the smallest stresses. monoMDS finds 2 dimensions and uses random configuration as the starting solution. The solution is iterative and there is no guaranteed convergence.

### 1.2.2 Overlay with Environment

See supporting text in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf):
  * 3 Environmental interpretation
  * 3.1 Vector fitting
  * 3.2 Surface fitting

```{r}
ef <- envfit(vare.mds, varechem, permu = 999)
ef
```

```{r}
plot(vare.mds, display = "sites")
plot(ef, p.max = 0.05)
```

**Question: What two soil chemistry elements have the strongest negative relationship with NMDS1 that is based on species composition?**

**Al and Fe have the strongest negative relationship with NMDS1 that is based on species composition.**

```{r}
ef <- envfit(vare.mds ~ Al + Ca, data = varechem)
plot(vare.mds, display = "sites")
plot(ef)

tmp <- with(varechem, ordisurf(vare.mds, Al, add = TRUE))
ordisurf(vare.mds ~ Ca, data=varechem, add = TRUE, col = "green4")
```
**The ordination surface plot from ordisurf() displays contours of an environmental gradient across sites. It is a more detailed look at an environmental gradient compared to the single blue line vector. This environmental overlay is generated by fitting a GAM where the response is the environmental variable of interest and the predictors are a bivariate smooth of the ordination axes, all given by the formula: Ca ~ s(NMDS1, NMDS2) (Remember each site is associated with a position on the NMDS axes and has an environmental value).  We can see from the code that the green4 color contours are for Calcium Ca. Describe in general terms (upper/lower/left/right/middle) where the highest and lowest values are found for Ca with respect to the ordination axes NMDS1 and NMDS2 (ie the ordination axes that describe the most variation of species composition between sites).**

**The highest values of Ca are found in the upper middle of the graph, as well as on the upper right. The Ca gradient is shown by the green contour lines and the highest value is 800 plotted on the upper middle, and similarly high values of 700 and 650 appear in the upper right. The lowest values of Ca are found in the lower left of the graph, with the lowest value being 300. Side note: the NMDS axes that plotted variation of species composition between sites used the Bray-Curtis method. The contour lines overlayed on top of the site points are considered to be a "fitted surface" of environmental variables. The direction of the Ca vector is pointing to the upper right because this is the direction of increase in this environmental variable across sites.**

### 1.2.3 Constrained Ordination on Species and Environment

See supporting text in [vegantutor.pdf](https://github.com/bbest/eds232-ml/raw/main/files/vegantutor.pdf):
  * 4 Constrained ordination
  * 4.1 Model specification

Technically, this uses another technique `cca`, or canonical correspondence analysis.

```{r}
# ordinate on species constrained by three soil elements
vare.cca <- cca(varespec ~ Al + P + K, varechem)
vare.cca
```
**Question: What is the difference between “constrained” versus “unconstrained” ordination within ecological context?**

**Unconstrained ordination can be considered indirect gradient analysis; the potential effects of factors that generated the output patterns can only be interpreted indirectly because these factors were not included in the analysis. Constrained ordination can be considered to be direct gradient analysis because it examines relationships between sets of variables. Constrained ordination includes 2+ sets of ecological information into the analysis, such as animals recorded as well as the environmental characteristics in the same data set.**

```{r}
# plot ordination
plot(vare.cca)
```

**Question: What sites are most differentiated by CCA1, i.e. furthest apart along its axis, based on species composition AND the environmnent? What is the strongest environmental vector for CCA1, i.e. longest environmental vector in the direction of the CCA1 axes?**

**The sites that are most differentiated by CCA1 are 28 and 4. The strongest environmental vector in the direction of the CCA1 axis is Al.**

```{r}
# plot 3 dimensions
ordiplot3d(vare.cca, type = "h")
```

```{r}
if (interactive()){
  ordirgl(vare.cca)
}
```






















